---
title: "R Markdown and Leaflet"
author: "Gerardo Mondragón"
date: "8/15/2020"
output: 
    html_document:
        keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readr)
library(leaflet)
library(dplyr)
```

## Overview

This web page is the assignment for week 2 of the Developing Data Products course. It was created using R Markdown and below is a map created with the **leaflet** package.

The map below will show the capitals of all countries in Europe, with the latitude and longitude taken from the following [site](http://techslides.com/list-of-countries-and-capitals). Then, we remove the capitals with less than 100,000 inhabitants, using data from [Wikipedia](https://en.wikipedia.org/wiki/List_of_national_capitals_by_population). 

Finally, they are plotted with the circles' radius being a function of the population for any given capital city. 

```{r read location data, warning=FALSE}
# read data
capitals <- read_csv("country-capitals.csv", 
    col_types = cols(CapitalLatitude = col_number(), 
        CapitalLongitude = col_number(), 
        CountryCode = col_skip()))

# subset Europe and delete extra columns
capitalsEurope <- capitals %>% filter(ContinentName == "Europe") %>% select(-5)
names(capitalsEurope) <- c("Country", "Capital", "Latitude", "Longitude")
```

```{r read population data, warning=FALSE}
population <- read_csv("capital-population.csv", 
    col_types = cols(Rank = col_skip(), `Country/Territory` = col_skip(), 
        Population = col_number(), Year = col_skip(), 
        `% ofcountry'spopulation` = col_skip()))

# merge
capitals2 <- merge(capitalsEurope, population, by="Capital")
capitals2 <- filter(capitals2, Population > 100000)
popup <- paste(capitals2$Capital, ",", capitals2$Country, "<br>", capitals2$Population)
## subset only lat and long
latlong <- capitals2 %>% select(c(3, 4))

```

```{r map}
# latlong %>% leaflet() %>% addTiles() %>% addMarkers(popup = popup, clusterOptions = markerClusterOptions())
latlong %>% 
    leaflet() %>% addTiles() %>% addCircles(weight = 2, radius = sqrt(capitals2$Population) * 100, popup = popup)

```
