---
title: "Reproducible Research Project 1"
author: "Gerardo Mondragón"
date: "8/2/2020"
output: html_document
---

## Setup
We load some packages to help us further down the line.
```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(ggplot2)
library(dplyr)
```

## Loading the data
Since the data is in a zip file, we unzip it first and then use **read.csv** on the output "activity.csv".
```{r zip}
unzip("activity.zip")
data <- read.csv("activity.csv")
head(data)
```

## What is the mean total number of steps taken per day?

### 0. Calculate the total number of steps taken per day
The data contains a lot of different readings for any given day. As such, we need to calculate the total number of steps per day. We do this using the **aggregate** function, grouping the **steps** component by **date**. We then rename the columns of this new data frame.
```{r steps per day}
steps_day <- aggregate(data$steps, list(data$date), sum)
names(steps_day) = c("date", "steps")
head(steps_day)
```

### 1. Make a histogram of the total number of steps taken each day
Using **dplyr** and **ggplot** we construct a histogram. We remove NA values, set a bin width of 1,000 and change some other formatting parameters.
```{r histogram}
steps_day %>% filter(!is.na(.$steps)) %>% ggplot(aes(steps)) + geom_histogram(binwidth = 1000, col = "black") + scale_x_continuous(breaks = seq(0,22000,2000)) + scale_y_continuous(breaks = seq(0,9,1)) + labs(x = "Steps", y = "Frequency", title = "Histogram of steps taken each day")
```

### 2. Calculate and report the mean and median total number of steps taken per day
This is a very straightforward calculation using the **mean** and **median** functions, whilst removing all NA values.
```{r mean and median}
mean(steps_day$steps, na.rm = T)
median(steps_day$steps, na.rm = T)
```

## What is the average daily activity pattern?

### 0. Create a table with the data
We need to create a table for the number of steps taken in each interval across all days. We use the **aggregate** function again.
```{r steps per time}
steps_time <- aggregate(steps~interval, data = data, FUN = mean)
head(steps_time)
```

### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
Again, we use the **ggplot** package to create the time series plot.
```{r time series}
ggplot(steps_time, aes(interval, steps)) + geom_line() + labs(x = "Interval", y = "Steps", title = "Average steps per interval")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
We use the **which.max** function to find the index of the maximum number of steps, and with this same index we find the interval to which it corresponds.
```{r max steps}
steps_time$interval[which.max(steps_time$steps)]
```

## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset
We use **sum** on **is.na** to count the number of NAs in the dataset.
```{r number of NAs}
sum(is.na(data$steps))
```

### 2. Devise a strategy for filling in all of the missing values in the dataset
We will use the rounded averages across all days for the interval to fill missing values. If the steps column is an NA, then it will replace it with the rounded average, otherwise it leaves the original value. 
```{r fill NAs}
data$complete <- ifelse(is.na(data$steps), round(steps_time$steps[match(data$interval, steps_time$interval)], 0), data$steps)
```

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r new dataset}
data_full <- data.frame(steps = data$complete, date = data$date, interval = data$interval)
head(data_full)
```

### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
As before, we use the **aggregate** function to sum all steps per day, rename the columns and then create a histogram using **ggplot**. The only difference in formatting is the upper limit of the y-axis. 
```{r steps per day full}
steps_day_full <- aggregate(data_full$steps, list(data_full$date), sum)
names(steps_day_full) = c("date", "steps")
ggplot(steps_day_full, aes(steps)) + geom_histogram(binwidth = 1000, col = "black") + scale_x_continuous(breaks = seq(0,22000,2000)) + scale_y_continuous(breaks = seq(0,20,1)) + labs(x = "Steps", y = "Frequency", title = "Histogram of steps taken each day")
```

Now we calculate the mean and median for this new dataset.
```{r new mean and median}
mean(steps_day_full$steps)
median(steps_day_full$steps)
```
We can observe that the difference between the original dataset and this new one with imputed data is very little.

## Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day
We first identify which day of the week it is. We use **as.Date** since the class of **date** is **chr**.
```{r day of the week}
data_full$day <- weekdays(as.Date(data_full$date))
```

Then we create the factor with the data from the code above.
```{r factor}
data_full$day_type <- as.factor(ifelse(data_full$day == "Saturday" | data_full$day == "Sunday", "weekend", "weekday"))
str(data_full)
```

### 2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
Once again, **aggregate** creates a table of the mean of steps grouped by interval and day type. Then **ggplot** and **facet_grid** take care of the rest. 
```{r panel time series}
steps_time_week <- aggregate(steps~interval+day_type, data = data_full, FUN = mean)
ggplot(steps_time_week, aes(interval, steps)) + geom_line() + labs(x = "Interval", y = "Steps", title = "Average steps per interval, weekdays vs weekends") + facet_grid(day_type~.)
```